Generating Random Logic Programs Using Constraint Programming

### 2020-08-13 seminar_talk/talk.tex
More slides

\documentclass{beamer}
\usepackage{tikz}
\usepackage[style=authoryear]{biblatex}
\usepackage{empheq}
\usepackage{pifont}
\usepackage{colortbl}
\usepackage{enumitem}

\setitemize{label=\usebeamerfont*{itemize item}%
  \usebeamercolor[fg]{itemize item}
  \usebeamertemplate{itemize item}}

@@

\begin{frame}{Probabilistic Inference}
  \begin{minipage}[t][0.5\textheight]{\textwidth}
    Let \alert{$a \oplus b \coloneqq a + b - ab$}. Then

@@

\begin{frame}{Inference Algorithms and Knowledge Compilation Maps}
  \begin{description}
  \item[NNF] negation normal form
  \item[BDD] binary decision diagrams
  \item[SDD] sentential decision diagrams
  \item[$k$-Best] only use the \alert{$k$} most probable proofs
  \item[d-DNNF] deterministic decomposable negation normal form
    \begin{itemize}
    \item<2-> for every pair \alert{$\alpha \lor \beta$}, we have \alert{$\alpha
        \land \beta = \bot$}
    \item<2-> for every pair \alert{$\alpha \land \beta$}, no atoms are shared
      between \alert{$\alpha$} and \alert{$\beta$}
    \item[\xmark\xmark]<3> $(A \lor C) \land (A \lor \neg B)$
    \item[\xmark\cmark]<3> $C \land (A \lor \neg B)$
    \item[\cmark\xmark]<3> $B \land C \land [(B \land A) \lor \neg B]$
    \item[\cmark\cmark]<3> $C \land [(B \land A) \lor \neg B]$
    \end{itemize}
  \end{description}
\end{frame}

@@

% TODO: a longer intro to experiments (the setup)
% TODO: conditional independence?
% TODO: and how is a predicate with its arguments represented?
% TODO: could add more experimental graphs
% TODO: a full(er) example of independence propagation
% TODO: could talk about what algorithms are behind the negative cycle detection
% TODO: for the bigraph talk, I had 28 slides full info and it still wasn't
% enough. Maybe I should aim for at least 30.
% TODO: more constraints
% TODO: more motivation and related work
\begin{frame}{Variable Symmetry Breaking}
  \begin{block}{The Problem}
    Let $\{ \textcolor{variable}{\mathsf{W}}, \textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}} \}$ be the set of variables. Then
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{Y}},
      \textcolor{variable}{\mathsf{X}})
    \]
    is equivalent to
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{X}},
      \textcolor{variable}{\mathsf{Y}})
    \]
    and to
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{W}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{X}},
      \textcolor{variable}{\mathsf{W}})
    \]
  \end{block}
\end{frame}

@@

% TODO: all the extra plots and example I had in the appendix
\begin{frame}{Stratification and Negative Cycles}
  \[
    \textcolor{probability}{0.1} \prob
    \textcolor{predicate}{\mathtt{friend}}(\textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}}) \ifff
    \negg\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}).
  \]
  \vfill
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{itemize}[leftmargin=0pt]
      \item<3-> \alert{Stratum 1}: predicates with no negation
      \item<3-> \alert{Stratum 2}: new predicates s.t. negative literals refer to
        predicates in \alert{Stratum~1}
      \end{itemize}
      \vspace{5em}
    \end{column}
    \begin{column}{0.6\textwidth}
      \hspace{-6.92em}
      \begin{tikzpicture}
        \node[draw,ellipse] (friend) {\textcolor{predicate}{friend}};
        \node[draw,ellipse,right=of friend] (influences) {\textcolor{predicate}{influences}};
        \node[draw,ellipse,below=of influences] (stress) {\textcolor{predicate}{stress}};
        \node[draw,ellipse,below=of stress] (spont) {\textcolor{predicate}{cancer\_spont}};
        \node[draw,ellipse,below=of spont] (smoke) {\textcolor{predicate}{cancer\_smoke}};
        \node[draw,ellipse,left=of spont] (person) {\textcolor{predicate}{person}};
        \node[draw,ellipse,right=of spont] (cancer) {\textcolor{predicate}{cancer}};
        \node[draw,ellipse,above=of cancer] (smokes) {\textcolor{predicate}{smokes}};
        \draw[-{Latex[length=2mm]}] (friend) -- (influences);
        \draw[-{Latex[length=2mm]}] (influences) -- (smokes);
        \draw[-{Latex[length=2mm]}] (smokes) -- (cancer);
        \draw[-{Latex[length=2mm]}] (stress) -- (smokes);
        \draw[-{Latex[length=2mm]}] (spont) -- (cancer);
        \draw[-{Latex[length=2mm]}] (smoke) -- (cancer);
        \draw[-{Latex[length=2mm]}] (person) -- (stress);
        \draw[-{Latex[length=2mm]}] (person) -- (spont);
        \draw[-{Latex[length=2mm]}] (person) -- (smoke);
        \draw<2->[-{Latex[length=2mm]},draw=red] (smokes) -- (friend);
      \end{tikzpicture}
    \end{column}
  \end{columns}
\end{frame}

@@

@@

@@

% * a full example for independence propagation (with an adjacency matrix)
% * specific constraints
% TODO: finish the slide
\begin{frame}{Summary}
  \begin{itemize}
  \item foo
  \item bar
  \item baz
  \end{itemize}
  \begin{block}{The implementation of the model is available at}
    \url{https://github.com/dilkas/random-logic-programs}
  \end{block}
\end{frame}

>>>>

\documentclass{beamer}
\usepackage{tikz}
\usepackage[style=authoryear]{biblatex}
\usepackage{empheq}
\usepackage{pifont}
\usepackage{colortbl}
\usepackage{soul}

@@

\begin{frame}{Probabilistic Inference: Reasoning by Hand}
  \begin{minipage}[t][0.5\textheight]{\textwidth}
    Let \alert{$a \oplus b \coloneqq a + b - ab$}. Then

@@

\begin{frame}{Probabilistic Inference: Probabilities of Worlds}
  \vspace{-0.5em}
  \onslide<2->{
    \[
      \textcolor{predicate}{\mathtt{cancer}}(\textcolor{constant}{\mathit{michelle}})
      = \alt<5>{\alert{\bot}}{\top}
    \]
    \[
      \Pr(\mathit{world}) =
      \only<2>{
        \textcolor{probability}{0.2} \times (1 - \textcolor{probability}{0.3})
        \times \textcolor{probability}{0.1} \times \textcolor{probability}{0.3}
      }
      \only<3>{
        \textcolor{probability}{0.2} \times \textcolor{probability}{0.3} \times
        \textcolor{probability}{0.1} \times (1 - \textcolor{probability}{0.3})
      }
      \only<4>{
        (1 - \textcolor{probability}{0.2}) \times (1 -
        \textcolor{probability}{0.3}) \times \textcolor{probability}{0.1} \times
        \textcolor{probability}{0.3}
      }
      \only<5>{
        (1 - \textcolor{probability}{0.2}) \times (1 -
        \textcolor{probability}{0.3}) \times (1 - \textcolor{probability}{0.1})
        \times (1 - \textcolor{probability}{0.3})
      }
    \]
  }
  \vspace{-1.5em}
  \begin{align*}
    \alt<2->{\hbox{\st{$\textcolor{probability}{0.2} \prob$}}}{\textcolor{probability}{0.2} \prob} &\alt<4,5>{\hbox{\st{$\textcolor{predicate}{\mathtt{stress}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{person}}(\textcolor{variable}{\mathsf{P}}).$}}}{\textcolor{predicate}{\mathtt{stress}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{person}}(\textcolor{variable}{\mathsf{P}}).} \\
    \alt<2->{\hbox{\st{$\textcolor{probability}{0.3} \prob$}}}{\textcolor{probability}{0.3} \prob} &\alt<2,4,5>{\hbox{\st{$\textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{P}_1}, \textcolor{variable}{\mathsf{P}_2}) \ifff \textcolor{predicate}{\mathtt{friend}}(\textcolor{variable}{\mathsf{P}_1}, \textcolor{variable}{\mathsf{P}_2}).$}}}{\textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{P}_1}, \textcolor{variable}{\mathsf{P}_2}) \ifff \textcolor{predicate}{\mathtt{friend}}(\textcolor{variable}{\mathsf{P}_1}, \textcolor{variable}{\mathsf{P}_2}).} \\
    \alt<2->{\hbox{\st{$\textcolor{probability}{0.1} \prob$}}}{\textcolor{probability}{0.1} \prob} &\alt<5>{\hbox{\st{$\textcolor{predicate}{\mathtt{cancer\_spont}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{person}}(\textcolor{variable}{\mathsf{P}}).$}}}{\textcolor{predicate}{\mathtt{cancer\_spont}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{person}}(\textcolor{variable}{\mathsf{P}}).} \\
    \alt<2->{\hbox{\st{$\textcolor{probability}{0.3} \prob$}}}{\textcolor{probability}{0.3} \prob} &\alt<3,5>{\hbox{\st{$\textcolor{predicate}{\mathtt{cancer\_smoke}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{person}}(\textcolor{variable}{\mathsf{P}}).$}}}{\textcolor{predicate}{\mathtt{cancer\_smoke}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{person}}(\textcolor{variable}{\mathsf{P}}).} \\
                                                                                                   &\alt<4,5>{\hbox{\st{$\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}) \ifff \textcolor{predicate}{\mathtt{stress}}(\textcolor{variable}{\mathsf{X}}).$}}}{\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}) \ifff \textcolor{predicate}{\mathtt{stress}}(\textcolor{variable}{\mathsf{X}}).} \\
                                                                                                   &\alt<2,4,5>{\hbox{\st{$\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}) \ifff \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}), \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{Y}}, \textcolor{variable}{\mathsf{X}}).$}}}{\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}) \ifff \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}), \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{Y}}, \textcolor{variable}{\mathsf{X}}).} \\
                                                                                                   &\alt<5>{\hbox{\st{$\textcolor{predicate}{\mathtt{cancer}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{cancer\_spont}}(\textcolor{variable}{\mathsf{P}}).$}}}{\textcolor{predicate}{\mathtt{cancer}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{cancer\_spont}}(\textcolor{variable}{\mathsf{P}}).} \\
                                                                                                   &\alt<3-5>{\hbox{\st{$\textcolor{predicate}{\mathtt{cancer}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{P}}), \textcolor{predicate}{\mathtt{cancer\_smoke}}(\textcolor{variable}{\mathsf{P}}).$}}}{\textcolor{predicate}{\mathtt{cancer}}(\textcolor{variable}{\mathsf{P}}) \ifff \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{P}}), \textcolor{predicate}{\mathtt{cancer\_smoke}}(\textcolor{variable}{\mathsf{P}}).} \\
                                                                                                   &\textcolor{predicate}{\mathtt{person}}(\textcolor{constant}{\mathit{michelle}}). \\
                                                                                                   &\textcolor{predicate}{\mathtt{person}}(\textcolor{constant}{\mathit{timothy}}). \\
                                                                                                   &\textcolor{predicate}{\mathtt{friend}}(\textcolor{constant}{\mathit{timothy}}, \textcolor{constant}{\mathit{michelle}}).
  \end{align*}
\end{frame}

\begin{frame}{Inference Algorithms and Knowledge Compilation Maps}
  \begin{description}
  \item[NNF] negation normal form
  \item[BDD] binary decision diagrams
  \item[SDD] sentential decision diagrams
  \item[$k$-Best] only use the \alert{$k$} most probable proofs
  \item[d-DNNF] deterministic decomposable negation normal form
    \begin{itemize}
    \item<2-> for every pair \alert{$\alpha \lor \beta$}, we have \alert{$\alpha
        \land \beta = \bot$}
    \item<2-> for every pair \alert{$\alpha \land \beta$}, no atoms are shared
      between \alert{$\alpha$} and \alert{$\beta$}
    \item<3-> examples:
      \begin{itemize}
      \item[\visible<4->{\xmark}\visible<5->{\xmark}]<3-> $(A \lor C) \land (A \lor \neg B)$
      \item[\visible<7->{\xmark}\visible<8->{\cmark}]<6-> $C \land (A \lor \neg B)$
      \item[\visible<10->{\cmark}\visible<11->{\xmark}]<9-> $B \land C \land [(B \land A) \lor \neg B]$
      \item[\visible<13->{\cmark}\visible<14->{\cmark}]<12-> $C \land [(B \land A) \lor \neg B]$
      \end{itemize}
    \end{itemize}
  \end{description}
\end{frame}

@@

% TODO: conditional independence?
% TODO: how is a predicate with its arguments represented?
% TODO: a full(er) example of independence propagation
% TODO: could talk about what algorithms are behind the negative cycle detection
% TODO: for the bigraph talk, I had 28 slides full info and it still wasn't
% enough. Maybe I should aim for at least 30. Right now I have 26.
% TODO: more constraints
% TODO: more motivation and related work
\begin{frame}{Variable Symmetry Breaking}
  \begin{block}{The Problem}
    Let $\{ \textcolor{variable}{\mathsf{W}}, \textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}} \}$ be the set of variables. Then
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{Y}},
      \textcolor{variable}{\mathsf{X}})
    \]
    is equivalent to
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{X}},
      \textcolor{variable}{\mathsf{Y}})
    \]
    and to
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{W}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{X}},
      \textcolor{variable}{\mathsf{W}})
    \]
  \end{block}
\end{frame}

@@

% TODO: all the extra plots and example I had in the appendix
\begin{frame}{Stratification and Negative Cycles}
  \[
    \textcolor{probability}{0.1} \prob
    \textcolor{predicate}{\mathtt{friend}}(\textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}}) \ifff
    \negg\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}).
  \]
  \vfill
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \setlength{\leftmargini}{0pt}
      \begin{itemize}
      \item<3-> \alert{Stratum 1}: predicates with no negation
      \item<3-> \alert{Stratum 2}: new predicates s.t. negative literals refer to
        predicates in \alert{Stratum~1}
      \end{itemize}
      \vspace{5em}
    \end{column}
    \begin{column}{0.6\textwidth}
      \hspace{-6.92em}
      \begin{tikzpicture}
        \node[draw,ellipse] (friend) {\textcolor{predicate}{friend}};
        \node[draw,ellipse,right=of friend] (influences) {\textcolor{predicate}{influences}};
        \node[draw,ellipse,below=of influences] (stress) {\textcolor{predicate}{stress}};
        \node[draw,ellipse,below=of stress] (spont) {\textcolor{predicate}{cancer\_spont}};
        \node[draw,ellipse,below=of spont] (smoke) {\textcolor{predicate}{cancer\_smoke}};
        \node[draw,ellipse,left=of spont] (person) {\textcolor{predicate}{person}};
        \node[draw,ellipse,right=of spont] (cancer) {\textcolor{predicate}{cancer}};
        \node[draw,ellipse,above=of cancer] (smokes) {\textcolor{predicate}{smokes}};
        \draw[-{Latex[length=2mm]}] (friend) -- (influences);
        \draw[-{Latex[length=2mm]}] (influences) -- (smokes);
        \draw[-{Latex[length=2mm]}] (smokes) -- (cancer);
        \draw[-{Latex[length=2mm]}] (stress) -- (smokes);
        \draw[-{Latex[length=2mm]}] (spont) -- (cancer);
        \draw[-{Latex[length=2mm]}] (smoke) -- (cancer);
        \draw[-{Latex[length=2mm]}] (person) -- (stress);
        \draw[-{Latex[length=2mm]}] (person) -- (spont);
        \draw[-{Latex[length=2mm]}] (person) -- (smoke);
        \draw<2->[-{Latex[length=2mm]},draw=red] (smokes) -- (friend);
      \end{tikzpicture}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{How Predicate Independence Affects Search Complexity}
  \centering
  \includegraphics{phase_transition.pdf}
\end{frame}

\begin{frame}{What Programs Should We Generate?}
  \begin{itemize}
  \item each program is divided into:
    \begin{itemize}
    \item rules
      \begin{itemize}
      \item e.g., $\textcolor{probability}{0.2} \prob
        \textcolor{predicate}{\mathtt{stress}}(\textcolor{variable}{\mathsf{P}})
        \ifff
        \textcolor{predicate}{\mathtt{person}}(\textcolor{variable}{\mathsf{P}}).$
      \end{itemize}
    \item facts
      \begin{itemize}
      \item e.g.,
        $\textcolor{predicate}{\mathtt{friend}}(\textcolor{constant}{\mathit{timothy}},
        \textcolor{constant}{\mathit{michelle}}).$
      \end{itemize}
    \end{itemize}
  \item predicates, variables, nodes: 2, 4, 8
  \item maximum arity: 1, 2, 3
  \item all possible numbers of pairs of independent predicates
  \item 10 programs per configuration
    \begin{itemize}
    \item fully restarting the constraint solver
    \end{itemize}
  \item probabilities sampled from $\{ 0.1, 0.2, \dots, 0.9 \}$
  \item query: random unlisted fact
  \end{itemize}
\end{frame}

\begin{frame}{Rules}
  \[
    \textcolor{probability}{0.2} \prob
    \textcolor{predicate}{\mathtt{stress}}(\textcolor{variable}{\mathsf{P}})
    \ifff
    \textcolor{predicate}{\mathtt{person}}(\textcolor{variable}{\mathsf{P}}).
  \]
  \bigskip
  \begin{itemize}
  \item no constants, no empty bodies
  \item one rule per predicate
  \item all rules are probabilistic
  \end{itemize}
\end{frame}

\begin{frame}{Facts}
  \[
    \textcolor{predicate}{\mathtt{friend}}(\textcolor{constant}{\mathit{timothy}},
    \textcolor{constant}{\mathit{michelle}}).
  \]
  \bigskip
  \begin{itemize}
  \item proportion probabilistic: 25\%, 50\%, 75\%
  \item constants: 100, 200, 400
  \item number of facts: $10^3$, $10^4$, $10^5$
    \begin{itemize}
    \item but only up to 75\% of all possible facts
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{How Encodings Compare Across Instances}
  \includegraphics{scatterplot.pdf}
\end{frame}

\begin{frame}{The Ratio of Listed Facts to Possible Facts}
  \includegraphics{proportion.pdf}
\end{frame}

@@

% TODO: finish the slide
\begin{frame}{Summary}
  \begin{itemize}
  \item foo
  \item bar
  \item baz
  \end{itemize}
  \begin{block}{The implementation of the model is available at}
    \url{https://github.com/dilkas/random-logic-programs}
  \end{block}
\end{frame}
